---
title: "ggplot2: Elegant Graphics for Data Analysis"
author: "Peter Baumgartner"
date: "2017/02/17"
output: 
  html_notebook: 
    pandoc_args: [ 
      "--number-offset=1,0" 
    ]
    fig_caption: true
    number_sections: yes
    toc: yes
    toc_depth: 3
---
```{r label = "global-options", echo=FALSE, highlight=TRUE}
knitr::opts_chunk$set(
        message = F,
        error = F,
        warning = F,
        comment = NA,
        highlight = T,
        prompt = T
        )
if (!(exists("tidyverse"))) {install.packages("tidyverse")}
        library(tidyverse)
# if (!(exists("oiblabs"))) {install.packages("oilabs")}
#        library(oilabs)
```
***

Working on the exercises of chapter 2 of [ggplot2: Elegant Graphics for Data] Analysis(https://www.amazon.de/ggplot2-Elegant-Graphics-Data-Analysis/dp/331924275X/ref=sr_1_1?ie=UTF8&qid=1487435428&sr=8-1&keywords=ggplot2), second edition. See also the material on GitHub [ggplot2-book](https://github.com/hadley/ggplot2-book), here especially [Chap.2](https://github.com/hadley/ggplot2-book/blob/master/ggplot.rmd).

>**Warning:** Keep in mind that I am still learning. There is no guarantee that my solutions are correct nor do they always present the most efficient way of doing things. Please feel free to comment on anything that you may think is wrong or somehow noteworthy (by writing an issue or by sending a pull request).

# Exercises of Chapter 2
## Exercises p.12/13
### How to get more info about the `mpg` dataset?
List five functions that you could use to get more information about the `mpg` dataset.

* `class()`     to see if it is a tibble and could be called without `head` or `tail` function
* `str()`       to look at the structure
* `glimpse()`   similiar but using the whole window width
* `head()`      first 6 record of the data set
* `tail()`      last 6 recors of the data set
* `View()`      load file into RStudio data viewer
* `summary()`   calculate summary statistics
```{r}
class(mpg)
str(mpg)
glimpse(mpg)
head(mpg)
tail(mpg)
names(mpg)
summary(mpg)
View(mpg)

```


### How can you find out what other datasets are included with ggplot2?
There is a listing on the special [documentation page](http://docs.ggplot2.org/current/index.html) for ggplot2 but there is also a general command `data()` for a listing of datasets for specific packages: More information with: `?data`
```{r list-available-data-sets}
data(package = "ggplot2")
?data
```

### How to convert fuel consumption into fuel economy?
Apart from the US, most countries use _fuel consumption_ (fuel consumed over fixed distance) rather than _fuel economy_ (distance travelled with fixed amount of fuel). How could you convert cty and hwy into the European standard of l/100km?

For the calcuation see the page on [Metric Conversions](http://www.metric-conversions.org): 

* 1 US liquid gallon = 3.78541 Litre
* 1 Mile = 1.609344 Kilometer

```{r}
mpgTol100km <- function(milespergallon){
  GalloLiter <- 3.785411784
  MileKilometer <- 1.609344 

  l100km <- round((100*GalloLiter)/(milespergallon*MileKilometer), digits = 1)
  l100km
}

mpg <- mpg %>%
        mutate(cty_1 = mpgTol100km(cty),
                  hwy_1 = mpgTol100km(hwy))
```

### How to get number of models and variations?
Which manufacturer has the most the models in this dataset? Which model has the most variations? Does your answer change if you remove the redundant specification of drive train (e.g. "pathfinder 4wd", "a4 quattro") from the model name?
```{r}
by_manufacturer <- group_by(mpg, manufacturer, sort = TRUE)
(models <- count(mpg, manufacturer, sort = TRUE))
by_models <- group_by(mpg, manufacturer, model)
per_model   <- summarise(by_models, nr = n())
(arrange(per_model, desc(nr)))
per_manufacturer <- summarise(per_model, "models" = n())
(arrange(per_manufacturer, desc(models)))
```
Dodge has the model variants and Toyota the most models. Model Dodge Caravan 2wd has the most variation.

### Changes if redundant specification of drive train is removed?

I do not understand the last part of the question: "Does your answer change if you remove the redundant specification of drive train (e.g. "pathfinder 4wd", "a4 quattro") from the model name?" 

What exactly is redundant? I could remove "quattro" from model "a4" from "audi" but this does not make sense with "pathfinder 4wd" from "nissan"

## Exercises p.14
### Describe the relationship between cty and hwy
How would you describe the relationship between cty and hwy? Do you have any concerns about drawing conclusions from that plot?

```{r}
ggplot(mpg, aes(cty, hwy)) +
        geom_point()

ggplot(mpg, aes(cty_1, hwy_1)) +
        geom_point()
```
There is a positive linear regression: More fuel in the city corresponds to more fuel on the highway, even if the consumption on the highway is on a somewhat lower level.

There is a lot of overplotting in this graph.

### What does the following plot show?
What does ggplot(mpg, aes(model, manufacturer)) + geom_point() show? Is it useful? How could you modify the data to make it more informative?

```{r}
ggplot(mpg, aes(model, manufacturer)) + 
        geom_point()

ggplot(mpg, aes(manufacturer, model)) + 
        geom_point()
```
The above graph shows the relationship between model and manufacturer. It is not useful as there is no relation between manufacturer and models. Every manufacturer is with its model separated and isolated, there is no connection or correlation. Moreover one cannot read the labels x axis and it would be ncessary to rotate the labels by 90 degrees.

[Manuel Steiner](https://github.com/M-E-Steiner/ggplot2-book-solutions/blob/master/Chapter2/Solutions_Chapter2.md#exercises-231-page-16) suggests to look at the total number of observations for each manufacturer-model combination using `geom_bar()`.
```{r}
df <- mpg %>% 
  transmute(manufacturer_model = paste(manufacturer, model, sep = " "))


ggplot(df, aes(manufacturer_model)) +
  geom_bar() + 
  coord_flip()
```
Maybe even better would be a graph with stacked bars, one bar for every manufacturer, divided by different colors representing the number of models of the manufacturer. But I do not know how to do this at the moment. (After I have learned more how to use `ggplot`, I will come back an try again.)

### Describe the following plots
Describe the data, aesthetic mappings and layers used for each of the following plots.
You'll need to guess a little because you haven't seen all the datasets and functions yet, but use your common sense! See if you can predict what the plot will look like before running the code.

* ggplot(mpg, aes(cty, hwy)) + geom_point()
* ggplot(diamonds, aes(carat, price)) + geom_point()
* ggplot(economics, aes(date, unemploy)) + geom_line()
* ggplot(mpg, aes(cty)) + geom_histogram()

```{r}
ggplot(mpg, aes(cty, hwy)) + geom_point()
ggplot(diamonds, aes(carat, price)) + geom_point()
ggplot(economics, aes(date, unemploy)) + geom_line()
ggplot(mpg, aes(cty)) + geom_histogram()
ggplot(mpg, aes(cty_1)) + geom_histogram()
```
Again [Manuel Steiner](https://github.com/M-E-Steiner/ggplot2-book-solutions/blob/master/Chapter2/Solutions_Chapter2.md#exercises-231-page-16) claims that there is help with the following summary function:
```{r}
summary(ggplot(economics, aes(date, unemploy)) + geom_line())
```
It seems to me that this is no desription or explanation of the resulting graphs.

* `ggplot(mpg, aes(cty, hwy)) + geom_point()`: It shows the graph already described.
* `ggplot(diamonds, aes(carat, price)) + geom_point()`: Relationship between weight (in carat) and price. It is interesting to observe two things: There is no linear relationship (after approximate 1 carat the prices rise extremly) and there are "vertical" lines. This second observation is the result of a cocentration of diamonds at the same rounded number of carats and the fact that diamonds with same carats have very different prices.
* `ggplot(economics, aes(date, unemploy)) + geom_line()`: This graph shows the rate of unemployment rate between 1967-07-01 and 2015-04-01. The rate is generally risng with a somewhat regular high approximately every 10 years.
* `ggplot(mpg, aes(cty)) + geom_histogram()`: This graph shows the fuel consumtion of the different car models. It can be seen that there the distribution is somewhat left skewed, whith a peak at 18 miles per gallon. 
* `ggplot(mpg, aes(cty_1)) + geom_histogram()`: I have added another graph, similar to the last one, but instead of miles per galoon (fuel consumption) this one shows the european equivalent of liter per kilometer (fuel economy). There is a similar left skewed distributen, but the appearance is very  dependent of the chosen number of bins.

## Exercises p.16
### Exerpiment with color, shape and size aesthetics

```{r}
vignette("ggplot2-specs")
ggplot(mpg, aes(displ, cty, color = class)) +
        geom_point()

ggplot(mpg, aes(displ, cty, color = "blue")) +
        geom_point()

ggplot(mpg, aes(displ, cty)) +
        geom_point(color = "blue")

ggplot(mpg, aes(displ, cty)) +
        geom_point(color = "blue", shape = 21, fill = "red", size = 3)

ggplot(mpg, aes(manufacturer, model)) +
        geom_point(color = "blue", shape = 18, fill = "red", size = 3) +
        geom_abline(slope = -0.0, intercept = 20, colour = "green", size = 1) +
        geom_vline(xintercept = 8, colour = "green", size = 1)

# Categorial
ggplot(mpg, aes(cty, displ, colour = class)) + 
  geom_point()

# Continuous 
ggplot(mpg, aes(cty, hwy, size = displ)) + 
  geom_jitter()

## Doesnt work for shape

# ggplot(mpg, aes(cty, hwy, shape = displ)) + 
#   geom_jitter()
```
### What happens if you map a continous variable to shape? Why?
```{r}
# ggplot(mpg, aes(cty, hwy, shape = displ)) +
#         geom_point()

```
It throws an error, because a continuous variable can not be mapped to shape. All aesthetics that have a natural continuous scale can be used for both continuous and discrete variables. Shape doesn't have a continuous scale so it throws an error. 

### What happens if you map trans to shape? Why?

```{r}
ggplot(mpg, aes(cty, hwy, color = trans)) +
        geom_point()

ggplot(mpg, aes(cty, hwy, shape = trans)) +
        geom_point()

```
	
> The shape palette can deal with a maximum of 6 discrete values because more than 6 becomes difficult to discriminate; you have 10. Consider specifying shapes manually if you must have them.
Removed 96 rows containing missing values (geom_point).

When a discrete variable has more than 6 different values its hard to discriminate hence we get a warning.

### How is drive train related to fuel economy? 
```{r}
ggplot(mpg, aes(cty, hwy, color = drv)) +
        geom_point()
```
The fuel economy of cars with front-wheel drive is generally better than of cars with 4 wd and rear wheel drive.

### How is drive train related to engine size and class?

```{r}
ggplot(mpg, aes(displ, class, color = drv)) +
        geom_point()
```
Cars with smaller engine size have front-wheel drive, pickup and suv have (with two exceptions in suv-clas) 4wd. 2seater and some of the subcompact cars have rear wheel drive.
