---
title: "06 Qualitative Data"
author: "Peter Baumgartner"
date: "2017-03-02"
output: 
        github_document:
                toc: true
                toc_depth: 4
        
---
***

```{r label = "global-options", echo=FALSE, highlight=TRUE}
knitr::opts_chunk$set(
        message = F,
        error = F,
        warning = F,
        comment = NA,
        highlight = T,
        prompt = T
        )
if (!require("tidyverse")) 
        {install.packages("tidyverse", repos = 'http://cran.wu.ac.at/')
        library(tidyverse)}
if (!require("reshape2"))
        {install.packages("reshape2", repos = 'http://cran.wu.ac.at/')
        library(reshape2)}
if (!require("pander"))
        {install.packages("pander", repos = 'http://cran.wu.ac.at/')
        library(pander)}

```


Yau, Chi. R Tutorial with Bayesian Statistics Using OpenBUGS (Kindle-Positionen524-525). Chi Yau. Kindle-Version. 

# Load data set into memory
```{r}
library(MASS)
head(painters)
```

# Frequency distribution

The frequency distribution of a data variable is a summary of the data occurrence in a collection of non-overlapping categories.

## Find the number of painters belonging to each school in the data set `painters`.

We apply the `table` function to compute the frequency distribution of the School variable. `table` uses the cross-classifying factors to build a contingency table of the counts at each combination of factor levels.

```{r frequency-distribution-1}
school <- painters$School
school.freq <- table(school)
school.freq
```

We can apply the `cbind`function to print the result in column format.
```{r result-in-table-format}
cbind(school.freq)
```
## Find the frequency distribution of the composition scores in `painters`

```{r frequency-distribution-2}
comp <- painters$Composition
comp.freq <- table(comp)
cbind(comp.freq)
```
## Find programmatically the school which has the most painters

```{r frequency-distribution-3-first-step}
school <- painters$School
school.freq <- table(school)
school.freq.max <- max(school.freq)
school.freq.max
```
```{r frequency-distribution-3-second-step}
L <- school.freq == school.freq.max
L
x <- school.freq[L]
x

```
An alternative way is to use the function `which`. Which returns the positions of all TRUE indices.
```{r frequency-distribution-3-second-step-alternativ}
y = which(school.freq == school.freq.max)
y
names(y)
```

Note that the obvious function which.max is insufficient. It only gives the first school that has the most painters, instead of _every_ school that has the most painters.

```{r}
which.max(school.freq)
```

# Relative frequencey distribution

In the data set `painters`, the relative frequency distribution of the School variable is a summary of the proportion of painters in each school.

## Find the proportion of painters distributed among the schools
```{r}
school <- painters$School
school.freq <- table(school)
school.freq.prop <- school.freq / nrow(painters)
school.freq.prop
```
We can print with fewer digits and make it more readable by setting the digits option. The options() function allows the user to set and examine a variety of global options which affect the way in which R computes and displays its results. In the case of digits is the standard value 7

```{r}
opt = options(digits = 1) # "It is a suggestion only."
school.freq.prop
options(opt)
```

In addition, we can apply the cbind function to print the percentage of painters in each school in column format. > old = options(digits=3)

```{r}
opt <- options(digits = 3)
cbind(school.freq.prop * 100)
options(opt)
```
## Find the relative frequency distribution of the composition scores

Find the relative frequency distribution of the composition scores in painters. 

* **Step 1:** Define a new variable comp that holds the composition scores. 
```{r}
comp = painters$Composition
```

* **Step 2:** Find the frequency distribution using the table function and save the result in another variable comp.freq
```{r}
comp.freq = table(comp)
```

* **Step 3:** Find the relative frequency distribution by dividing comp.freq with the sample size. 
```{r}
comp.rel.prop = comp.freq / nrow(painters)
comp.rel.prop
```


```{r}
old = options(digits = 3)
cbind(comp.rel.prop * 100)
```

# Bar Graph

## Find the bar graph of the painter schools in the data set `painters
```{r drawing-bar-graph}
school <- painters$School
school.freq <- table(school)
barplot(school.freq)
```

