---
title: "Basic Datasets in R"
author: "Peter Baumgartner"
date: "Jan 18, 2017"
output:
  html_notebook:
    toc: yes
---
        
```{r label = "global-options", echo=FALSE}
knitr::opts_chunk$set(message = F, error = F, warning = F, comment = NA)
```
***
Here I am experimenting with "getting & cleaning data". My goal is to load the basic R datasets as tidy data (each row refers to a dataset with two variables: Name and Description).

# Getting data

Load the list of all datasets into a Variable
```{r}
ds <- library(help = "datasets")
```
Inspect the structure of the list
```{r}
str(ds)
```
str(ds) shows that the interesting part for me is the character vector ds$info[[2]]


# Tidy data

The interesting data are now in R. The next step is to clean it, so that it forms a tidy table. 
```{r}
dStr <- ds$info[[2]]
cat("Length of character vector =", length(dStr), "\n\n")
cat("The first five strings:\n")
head(dStr, 5)
cat("\n\nThe last five strings:\n")
tail(dStr, 5)
```
After detailed inspecting it turned outm, that some of the exact hundert strings are continuation, as the descriptions did not fit in one line of the packageInfo format.

To find these strings I try to use a regular expression:
```{r}

i <- grep("^ +", dStr) # returns a vector of the indices of the elements of dStr that yielded a match
for (val in i) {
        dStr[val - 1] <- paste(dStr[val - 1], str_trim(dStr[val]))
        dStr[val] <- "" # do not remove line because that would scramble the index i 
}
nStr <- dStr[dStr != ""]
```


This procedure has several phases which I embed in the function of vapply.

1. Split the vector string after two spaces to get Name & Description
2. Name the columns. 
3. Naming the columns tells vapply tp produce an array (and not an vector)
4. Transpose (t) the matrix = change rows and columns
```{r}
library("stringr")
df <- t(vapply(
        nStr,
        str_split_fixed,
        c("Name" = "", "Description" = ""),
        pattern = " ",
        n = 2,
        USE.NAMES = FALSE
))
```

