---
title: "Subsetting and Sorting"
output:
  html_notebook:
    toc: yes
---
```{r label = "global-options", echo=FALSE}
knitr::opts_chunk$set(message = F, error = F, warning = F, comment = NA)
```

## Subsetting - a quick review
### Creating a data frame to work with
```{r label = "generate-df"}
set.seed(13435)
X <-
        data.frame(
        "var1" = sample(1:5),
        "var2" = sample(6:10),
        "var3" = sample(11:15)
        )
X <- X[sample(1:5), ]
X$var2[c(1, 3)] = NA
X
```

### Get the first column
```{r label = "get-first-column"}
X[, 1]
```
### Get the first row
```{r label = "get-first-row"}

X[1, ]
```
### Get the first variable in the first row and first column
```{r label = get-first-var-first-row-first-col}
X[1, 1]
```
### Get (row) variable by name
```{r label = "get-var-by-name"}
X[, "var2"]
X[2, "var2"]
```
### Get for a row subset a variable by name
```{r label = "get-row-subset-and-var-by-name"}
X[1:2, "var2"]
```
### Get data with logical AND conditions
```{r label = "get-data-with-logical-and-condition"}
X[(X$var1 >= 3 & X$var3 >= 13), ]
```
### Get data with logical OR conditions
```{r label = "get-data-with-logical-or-condition"}
X[(X$var1 >= 3 | X$var3 >= 13), ]
```

## Dealing with missing values
### Problem: Subsetting on NAs
Subsetting on NAs will not produce the actual rows:
```{r label = "na-subsetting-problem"}
X[(X$var2 > 8),]
```

### Solution: Subsetting with `which`
You can use the behavior of the `which` function, which takes only TRUE indices. 
```{r label = use-which-with-na-subsetting-problem}
X[which(X$var2 > 8),]
```

## Sorting, ordering, arrange of rows
### Sort: Ordering of just one variable
```{r label = "order-one-var"}
sort(X$var1)
sort(X$var1, decreasing = TRUE)
sort(X$var2) # NAs are not included
sort(X$var2, na.last = TRUE)

```
### Order: Ordering of a row in a data frame
```{r label = "order-df"}
X[order(X$var1),]
```
### Order: Ordering of multiple variables of a data frame
The following example is not a good one, as there aren't any ties where you can see that the second variable is ordered inside the first
```{r label = "order-multiple-vars-in-df"}
X[order(X$var1,X$var3),] 
```
### Arrange: Ordering with the arrange command of `dplyr`
```{r label = "order-with-arrange"}`
library("dplyr")
arrange(X, var1)
### Arrange: Sorting in descending order
```
```{r label = "order-desc-with-arrange"}
library("dplyr")
arrange(X, desc(var1))
```

## Adding row and columns
### Adding directly at the end of the data frame 
```{r label = "add-col-directly"}
X$var4 <- rnorm(5)
X
```
### Add column to the right with cbind
```{r label = "add-col-right-with-cbind"}
Yr <- cbind(X,rnorm(5)) # it has to be the same dimension as the data frame
Yr
```
### Add column to the left with cbind
```{r label = "add-col-left-with-cbind"}
Yl <- cbind(rnorm(5), X) # it has to be the same dimension as the data frame
Yl
```
### Add row at the end with rbind
```{r label = "add-row-below-with-rbind"}
Xr <- rbind(X, c(0, 4, 22, NA)) # it has to be the same dimension as the data frame
Xr
```
### Add row to the top with rbind
```{r label = "add-row-above-with-rbind"}

Yl <- rbind(c(0, 4, 22, NA), X) # it has to be the same dimension as the data frame
Yl
```

## Subsetting with lists
Lists are a very generic datatype. It can hold vectors, strings, matrices, models, list of other list. Lists can reference data using $ (if the elements are named), or using [], or [[]]. For instance, if there is a list "mylist", with a list of "myname" then:

1. [], mylist["myname"]  returns a list of elements
2. [[]], mylist$myname, mylist[["myname"]] returns the original class (vector, matrix etc.)

The most important distinction between [, [[ and $ is that the [ can select more than one element whereas the other two select a single element.
### Creating a list to work with
```{r label = "creating-a-list"}
mylist <-
        list(letters = c("A", "b", "c"),
        numbers = 1:3,
        matrix(1:25, ncol = 5),
        text = list(text1 = c("This", "is", "a", "test"), 
             text2 = c("That", "is", "another", "test")))
mylist
```
### There are two different forms of subsetting with lists
#### Mode 1: Subsetting returns a list
```{r label = "subsetting-returns-list"}
mylist[1] # returns a list
class(mylist[1])
mylist["letters"] # returns a list
class(mylist["letters"])
```
#### Mode 2: Subsetting returns data of the class of the subsetted datatype
```{r label = "subsetting-returns-vector"}
mylist[[1]] # returns the vector 'letters'
class(mylist[[1]])
mylist$letters # returns vector
class(mylist$letters)
mylist[["letters"]] # returns the vector 'letters
class(mylist[["letters"]])
```
