---
title: "Differences between knit and run code in learnr tutorials"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Goal of the Exerpiment

I want to see the differences between `knit` and `Run All` code.

### Setup

I will create two chunks:

1. The first one creates a variable `var1` and display it.
2. The second one uses the variable `var1`, creates a second variable `var2` and displays it.

### Procedures

1. Knit document without ever running any chunks.
2. Run first chunk and knit document.
3. Run second chunk and knit document.
4. Change first chunk and knit document.
5. Change first chunk again and "Run All".


```{r var1}
var1 <- 1 + 1
var1
```

```{r var2}
var2 <- var1 + 10
var2
```

### Results

As expected: There is **no difference**. knit (Run Document) creates a new (empty) environment, runs all chunks and knit them.

### Next question

How about setup chunks in learner exercises?. What is their behavior?
